Biogeography \& environmental conditions shape bacteriophage-bacteria networks across the human microbiome

### 2018-03-07 doc/manuscript.tex
updated with pats edits and fixed a bug with edge count near beginning.

The human microbiome, the collection of microbial communities that
colonize the human body, is a crucial component to health and disease.
Two major components to the human microbiome are the bacterial and viral
communities. These communities have primarily been studied separately
using metrics of community composition and diversity. These approaches
have failed to capture the complex dynamics of interacting bacteria and
phage communities, which frequently share genetic information and work
together to maintain ecosystem homestatsis (e.g.~kill-the-winner
dynamics). Removal of bacteria or phage can disrupt or even collapse
those ecosystems. Relationship-based network approaches allow us to
capture this interaction information. Using this network-based approach
with three independent human cohorts, we were able to present an initial
understanding of how phage-bacteria networks differ throughout the human
body, so as to provide a baseline for future studies of how and why
microbiome networks differ in disease states.

@@

The bacterial and viral sequences from these studies were quality
filtered and assembled into contigs (``genomic fragments''). We further
grouped the related bacterial and phage contigs into operationally
defined units based on their k-mer frequencies and co-abundance
patterns, similar to previous reports \textbf{(Supplemental Figure
\ref{ContigStats} - \ref{ClusterStats})} {[}42{]}. This was done both
for dimensionality reduction and to prevent inflation of node counts by
using contigs which are expected to represent multiple fragments from
the same genomes. This was also done to create genome analogs that we
could use in our classification model which was built using genome
sequences. We referred to these operationally defined groups of related
contigs as operational genomic units (OGUs). Each OGU represented a
genomically similar sub-population of either bacteria or phages. Contig
lengths within clusters ranged between \(10^{3}\) and \(10^{5.5}\) bp
\textbf{(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})}.

The original publications reported that the whole metagenomic shotgun
sequence samples, which primarily consisted of bacteria, had an average
viral relative abundance of 0.4\% \textbf{(Table S1)} {[}13,14,38,39{]}.
We confirmed these reports by finding that only 2\% (6 / 280 OGUs) of
bacterial OGUs had significantly strong nucleotide similarity to phage
reference genomes (e-value \textless{} \(10^{-25}\)) {[}13,14,38,39{]}.
Additionally, no OGUs were confidently identified as lytic or temperate
phage OGUs in the bacterial dataset using the Virsorter algorithm
{[}50{]}.

We also supplemented the previous virome fraction quality control
measures \textbf{(Table S1)} to find that, in light of the rigorous
purification and quality control during sample collection and
preparation, 77\% (228 / 298 operational genomic units) still had some
nucleotide similarity to a given bacterial reference genome (e-value
\textless{} \(10^{-25}\)). It is important to note that interpreting
such alignment is complicated by the fact that most reference bacterial
genomes also contain prophages (phages integrated into bacterial
genomes), meaning we do not know to what extent the alignments were the
result of bacterial contaminants in the virome fraction and what were
true integrated prophages. As most phages in these communities have been
shown to be temperate (they integrate into bacterial genomes) using
methods that included nucleotide alignments of phages to bacterial
reference genomes, we expected that a large fraction of those phages
were temperate and therefore shared elements with bacterial reference
genomes -- a trend previously reported {[}14{]}. To ensure the purity of
our sample sets, we supplemented the quality control measures by
filtering out all OGUs that could be potential bacterial contaminants,
as described previously {[}42{]}. This resulted in the removal of 143
OGUs that exhibited nucleotide similarity to bacterial genomes but no
identifiable known phage elements. We were also able to identify two
OGUs as representing \textbf{complete}, high confidence phages using the
stringent Virsorter phage identification algorithm (class 1 confidence
group) {[}50{]}.

@@

We predicted which phage OGUs infected which bacterial OGUs using a
random forest model trained on experimentally validated infectious
relationships from six previous publications {[}41,51--55{]}. Only
bacteria and bacteriophages were used in the model. The training set
contained 43 diverse bacterial species and 30 diverse phage strains,
including both broad and specific ranges of infection
\textbf{(Supplemental Figure \ref{ValidationOverview} A - B, Table S2)}.
While it is true that there are more known phages that infect bacteria,
we were limited by the information confirming which phages do not infect
certain bacteria and attempted to keep the numbers of positive and
negative interactions similar. Phages with linear and circular genomes,
as well as ssDNA and dsDNA genomes, were included in the analysis.
Because we used DNA sequencing studies, RNA phages were not considered
\textbf{(Supplemental Figure \ref{ValidationOverview} C-D)}. This
training set included both positive relationships (a phage infects a
bacterium) and negative relationships (a phage does not infect a
bacterium). This allowed us to validate the false positive and false
negative rates associated with our candidate models, thereby building
upon previous work that only considered positive relationships {[}41{]}.
It is also worth noting that while a positive interaction is strong
evidence that the interaction exists, we must also be conscious that
negative interactions are only weak evidence for a lack of interaction
because the finding could be the result of our inability to reproduce
conditions in which those interactions occur. Altogether we decided to
maintain a balanced dataset at the cost of under-sampling the available
positive interaction information because the use of such a severely
unbalanced dataset often results in over-fit and uninformative model
training. However, as an additional validation measure, we used the
extensive additional positive interactions as a secondary dataset to
confirm that we could identify infectious interactions from a more
diverse bacterial and phage dataset. Using this approach, we confirmed
that 382 additional phage reference genomes, representing a diverse
range of phages, were matched to at least one reference bacterial host
genome of the species that they were expected to infect
\textbf{(Supplemental Figure \ref{secondval})}. Because the model was
built on full genomes and used on OGUs, we also assessed whether our
model was resilient to incomplete reference genomes. We found that the
use of our model on random contigs representing as little as 50\% length
of the original reference phage and bacterial genomes resulted in
minimal reduction in the ability of the model to identify relationships
\textbf{(Supplemental Figure \ref{contigval})}.

Four phage and bacterial genomic features were used in our random forest
model to predict infectious relationships between bacteria and phages:
1) genome nucleotide similarities, 2) gene amino acid sequence
similarities, 3) bacterial Clustered Regularly Interspaced Short
Palindromic Repeat (CRISPR) spacer sequences that target phages, and 4)
similarity of protein families associated with experimentally identified
protein-protein interactions {[}56{]}. These features were calculated
using the training set described above. While the nucleotide and amino
acid similarity metrics are expected to identify prophage signatures,
the protein family interaction and CRISPR signatures are expected to aid
in identifying lytic phages in addition to temperate phages. We chose to
utilize these metrics that directly compare nucleotide sequences between
sample phages and bacteria, instead of relying on alignment to reference
genomes or known marker genes, because we are extrapolating our model to
highly diverse communities which we expect to diverge significantly from
the available reference genomes. The resulting random forest model was
assessed using nested cross validation, and the median area under its
receiver operating characteristic (ROC) curve was 0.788, the median
model sensitivity was 0.905, and median specificity was 0.538
\textbf{(Figure \ref{RocCurve} A)}. This balance of confident true
positives at the cost of fewer true negatives is ideal for this type of
dataset which consists of primarily positive connections
\textbf{(Supplemental Figure \ref{nestedperformance})}. Nested cross
validation of the model demonstrated that the sensitivity and
specificity of the model could vary but the overall model performance
(AUC) remained more consistent \textbf{(Supplemental Figure
\ref{crossval})}. This suggested that our model would perform with a
similar overall accuracy despite changes in sensitivity/specificity
trade-offs. The most important predictor in the model was amino acid
similarity between genes, followed by nucleotide similarity of whole
genomes \textbf{(Figure \ref{RocCurve} B)}. Protein family interactions
were moderately important to the model, and CRISPRs were largely
uninformative, due to the minimal amount of identifiable CRISPRs in the
dataset and their redundancy with the nucleotide similarity methods
\textbf{(Figure \ref{RocCurve} B)}. Approximately one third of the
training set relationships yielded no score and therefore were unable to
be assigned an interaction prediction \textbf{(Figure \ref{RocCurve}
C)}.

@@

We used our random forest model to classify the relationships between
bacteria and phage operational genomic units, which were then used to
build the interactive network. The master network, analogous to the
universal microbiome network concept previously described {[}57{]},
contained the three studies as sub-networks, which themselves each
contained sub-networks for each sample \textbf{(Supplemental Figure
\ref{NetworkDiagram})}. Metadata including study, sample ID, disease,
and OGU abundance within the community were stored in the master network
for parsing in downstream analyses \textbf{(Supplemental Figure
\ref{NetworkDiagram})}. The phage and bacteria of the master network
demonstrated both narrow broad ranges of infectious interactions
\textbf{(Supplemental Figure \ref{inthm})}. Bacterial and phage relative
abundance was recorded in each sample for each OGU and the weight of the
edge connecting those OGUs was calculated as a function of those
relative abundance values. The separate extraction of the phage and
bacterial libraries ensured a more accurate measurement of the microbial
communities, as has been outlined previously {[}58,59{]}. The master
network was highly connected and contained 38,337 infectious
relationships among 435 nodes, representing 155 phages and 280 bacteria.
Although the network was highly connected, not all relationships were
present in all samples. Relationships were weighted by the relative
abundances of their associated bacteria and phages. Like the master
network, the skin network exhibited a diameter of 4 (measure of graph
size; the greatest number of traversed vertices required between two
vertices) and included 433 ( phages, bacteria, 99.5\% total) and 38,099
(99.4\%) of the master network nodes and edges, respectively
\textbf{(Figure \ref{RocCurve} E - F)}. Additionally, the subnetworks
demonstrated narrow ranges of eccentricity across their nodes
\textbf{(Supplemental Figure \ref{eccplot})}. The phages and bacteria in
the diet and twin sample sets were more sparsely related, with the diet
study consisting of 80 ( phages, bacteria) nodes and 4,692
relationships, and the twin study containing 130 ( phages, bacteria)
nodes and 14,684 relationships \textbf{(Figure \ref{RocCurve} E - F)}.
As an interesting validation measure, we identified five (1.7\%)
examples of phage OGUs which contained similar genomic elements to the
four previously described, broadly infectious phages isolated from Lake
Michigan (tblastx; e-value \textless{} \(10^{-25}\)) {[}60{]}.

>>>>

The human microbiome, the collection of microbial communities that
colonize the human body, is a crucial component to health and disease.
Two major components of the human microbiome are the bacterial and viral
communities. These communities have primarily been studied separately
using metrics of community composition and diversity. These approaches
have failed to capture the complex dynamics of interacting bacteria and
phage communities, which frequently share genetic information and work
together to maintain ecosystem homestatsis (e.g.~kill-the-winner
dynamics). Removal of bacteria or phage can disrupt or even collapse
those ecosystems. Relationship-based network approaches allow us to
capture this interaction information. Using this network-based approach
with three independent human cohorts, we were able to present an initial
understanding of how phage-bacteria networks differ throughout the human
body, so as to provide a baseline for future studies of how and why
microbiome networks differ in disease states.

@@

The bacterial and viral sequences from these studies were quality
filtered and assembled into contigs (i.e.~genomic fragments). We further
grouped the related bacterial and phage contigs into operationally
defined units based on their k-mer frequencies and co-abundance
patterns, similar to previous reports \textbf{(Supplemental Figure
\ref{ContigStats} - \ref{ClusterStats})} {[}42{]}. This was done both
for dimensionality reduction and to prevent inflation of node counts by
using contigs which are expected to represent multiple fragments from
the same genomes. This was also done to create genome analogs that we
could use in our classification model which was built using genome
sequences. We referred to these operationally defined groups of related
contigs as operational genomic units (OGUs). Each OGU represented a
genomically similar sub-population of either bacteria or phages. Contig
lengths within clusters ranged between \(10^{3}\) and \(10^{5.5}\) bp
\textbf{(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})}.

The original publications reported that the whole metagenomic shotgun
sequence samples, which primarily consisted of bacteria, had an average
viral relative abundance of 0.4\% \textbf{(Table S1)} {[}13,14,38,39{]}.
We confirmed these reports by finding that only 2\% (6 / 280 OGUs) of
bacterial OGUs had significantly strong nucleotide similarity to phage
reference genomes (e-value \textless{} \(10^{-25}\)) {[}13,14,38,39{]}.
Additionally, no OGUs were confidently identified as lytic or temperate
phage OGUs in the bacterial dataset using the Virsorter algorithm
{[}50{]}. We also supplemented the previous virome fraction quality
control measures \textbf{(Table S1)} to find that, in light of the
rigorous purification and quality control during sample collection and
preparation, 77\% (228 / 298 operational genomic units) still had some
nucleotide similarity to a given bacterial reference genome (e-value
\textless{} \(10^{-25}\)). It is important to note that interpreting
such alignment is complicated by the fact that most reference bacterial
genomes also contain prophages (i.e.~phages integrated into bacterial
genomes), meaning we do not know to what extent the alignments were the
result of bacterial contaminants in the virome fraction and what were
true integrated prophages. As most phages in these communities have been
shown to be temperate (i.e.~they integrate into bacterial genomes) using
methods that included nucleotide alignments of phages to bacterial
reference genomes, we expected that a large fraction of those phages
were temperate and therefore shared elements with bacterial reference
genomes -- a trend previously reported {[}14{]}. To ensure the purity of
our sample sets, we supplemented the quality control measures by
filtering out all OGUs that could be potential bacterial contaminants,
as described previously {[}42{]}. This resulted in the removal of 143
OGUs that exhibited nucleotide similarity to bacterial genomes but no
identifiable known phage elements. We were also able to identify two
OGUs as representing \textbf{complete}, high confidence phages using the
stringent Virsorter phage identification algorithm (class 1 confidence
group) {[}50{]}.

@@

We predicted which phage OGUs infected which bacterial OGUs using a
random forest model trained on experimentally validated infectious
relationships from six previous publications {[}41,51--55{]}. Only
bacteria and bacteriophages were used in the model. The training set
contained 43 diverse bacterial species and 30 diverse phage strains,
including both broad and specific ranges of infection
\textbf{(Supplemental Figure \ref{ValidationOverview} A - B, Table S2)}.
While it is true that there are more known phages that infect bacteria,
we were limited by the information confirming which phages do not infect
certain bacteria and attempted to keep the numbers of positive and
negative interactions similar. Phages with linear and circular genomes,
as well as ssDNA and dsDNA genomes, were included in the analysis.
Because we used DNA sequencing studies, RNA phages were not considered
\textbf{(Supplemental Figure \ref{ValidationOverview} C-D)}. This
training set included both positive relationships (i.e.~a phage infects
a bacterium) and negative relationships (i.e.~a phage does not infect a
bacterium). This allowed us to validate the false positive and false
negative rates associated with our candidate models, thereby building
upon previous work that only considered positive relationships {[}41{]}.
It is also worth noting that while a positive interaction is strong
evidence that the interaction exists, we must also be conscious that
negative interactions are only weak evidence for a lack of interaction
because the finding could be the result of our inability to reproduce
conditions in which those interactions occur. Altogether we decided to
maintain a balanced dataset at the cost of under-sampling the available
positive interaction information because the use of such a severely
unbalanced dataset often results in over-fit and uninformative model
training. However, as an additional validation measure, we used the
extensive additional positive interactions as a secondary dataset to
confirm that we could identify infectious interactions from a more
diverse bacterial and phage dataset. Using this approach, we confirmed
that 382 additional phage reference genomes, representing a diverse
range of phages, were matched to at least one reference bacterial host
genome of the species that they were expected to infect
\textbf{(Supplemental Figure \ref{secondval})}. Because the model was
built on full genomes and used on OGUs, we also assessed whether our
model was resilient to incomplete reference genomes. We found that the
use of our model on random contigs representing as little as 50\% length
of the original reference phage and bacterial genomes resulted in
minimal reduction in the ability of the model to identify relationships
\textbf{(Supplemental Figure \ref{contigval})}.

Four phage and bacterial genomic features were used in our random forest
model to predict infectious relationships between bacteria and phages:
1) genome nucleotide similarities, 2) gene amino acid sequence
similarities, 3) bacterial Clustered Regularly Interspaced Short
Palindromic Repeat (CRISPR) spacer sequences that target phages, and 4)
similarity of protein families associated with experimentally identified
protein-protein interactions {[}56{]}. These features were calculated
using the training set described above. While the nucleotide and amino
acid similarity metrics were expected to identify prophage signatures,
the protein family interaction and CRISPR signatures were expected to
aid in identifying lytic phages in addition to temperate phages. We
chose to utilize these metrics that directly compare nucleotide
sequences between sample phages and bacteria, instead of relying on
alignment to reference genomes or known marker genes, because we were
extrapolating our model to highly diverse communities which we expect to
diverge significantly from the available reference genomes. The
resulting random forest model was assessed using nested cross
validation, and the median area under its receiver operating
characteristic (ROC) curve was 0.788, the median model sensitivity was
0.905, and median specificity was 0.538 \textbf{(Figure \ref{RocCurve}
A)}. This balance of confident true positives at the cost of fewer true
negatives was ideal for this type of dataset which consisted of
primarily positive connections \textbf{(Supplemental Figure
\ref{nestedperformance})}. Nested cross validation of the model
demonstrated that the sensitivity and specificity of the model could
vary but the overall model performance (AUC) remained more consistent
\textbf{(Supplemental Figure \ref{crossval})}. This suggested that our
model would perform with a similar overall accuracy despite changes in
sensitivity/specificity trade-offs. The most important predictor in the
model was amino acid similarity between genes, followed by nucleotide
similarity of whole genomes \textbf{(Figure \ref{RocCurve} B)}. Protein
family interactions were moderately important to the model, and CRISPRs
were largely uninformative, due to the minimal amount of identifiable
CRISPRs in the dataset and their redundancy with the nucleotide
similarity methods \textbf{(Figure \ref{RocCurve} B)}. Approximately one
third of the training set relationships yielded no score and therefore
were unable to be assigned an interaction prediction \textbf{(Figure
\ref{RocCurve} C)}.

@@

We used our random forest model to classify the relationships between
bacteria and phage operational genomic units, which were then used to
build the interactive network. The master network, analogous to the
universal microbiome network concept previously described {[}57{]},
contained the three studies as sub-networks, which themselves each
contained sub-networks for each sample \textbf{(Supplemental Figure
\ref{NetworkDiagram})}. Metadata including study, sample ID, disease,
and OGU abundance within the community were stored in the master network
for parsing in downstream analyses \textbf{(Supplemental Figure
\ref{NetworkDiagram})}. The phage and bacteria of the master network
demonstrated both narrow broad ranges of infectious interactions
\textbf{(Supplemental Figure \ref{inthm})}. Bacterial and phage relative
abundance was recorded in each sample for each OGU and the weight of the
edge connecting those OGUs was calculated as a function of those
relative abundance values. The separate extraction of the phage and
bacterial libraries ensured a more accurate measurement of the microbial
communities, as has been outlined previously {[}58,59{]}. The master
network was highly connected and contained 38,337 infectious
relationships among 435 nodes, representing 155 phages and 280 bacteria.
Although the network was highly connected, not all relationships were
present in all samples. Relationships were weighted by the relative
abundances of their associated bacteria and phages. Like the master
network, the skin network exhibited a diameter of 4 (measure of graph
size; the greatest number of traversed vertices required between two
vertices) and included 433 (154 phages, 279 bacteria, 99.5\% total) and
38,099 (99.4\%) of the master network nodes and edges, respectively
\textbf{(Figure \ref{RocCurve} E - F)}. Additionally, the subnetworks
demonstrated narrow ranges of eccentricity across their nodes
\textbf{(Supplemental Figure \ref{eccplot})}. The phages and bacteria in
the diet and twin sample sets were more sparsely related, with the diet
study consisting of 80 (32 phages, 48 bacteria) nodes and 1,290
relationships, and the twin study containing 130 (29 phages, 101
bacteria) nodes and 2,457 relationships \textbf{(Figure \ref{RocCurve} E
- F)}. As a validation measure, we identified five (1.7\%) examples of
phage OGUs which contained similar genomic elements to the four
previously described, broadly infectious phages isolated from Lake
Michigan (tblastx; e-value \textless{} \(10^{-25}\)) {[}60{]}.
